---
title: "Binary Mixtures of Linear Models"
author: "S Wotherspoon"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Binary Mixtures of Linear Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,fig.width=6,fig.height=5)
```


## Model

The `bmixlm` model assumes that there are two linear models
\[
\begin{align}
    Y_{1} &\sim \operatorname{N}(X_{1} \beta_{1}, \sigma_{1}^{2})\\
    Y_{2} &\sim \operatorname{N}(X_{2} \beta_{2}, \sigma_{2}^{2})
\end{align}
\]
and a set of latent binary indicators governed by a probit model
\[
\begin{align}
    b &\sim \operatorname{Bin}(1,p)\\
    \Phi^{-1}(p) &= X_{p} \beta_{p}
\end{align}
\]
where the design matrices \(X_{1}\), \(X_{2}\) and \(X_{p}\) are
generated from a common set of covariates \(x_{1},
x_{2},\ldots,x_{n}\), and the observed response \(y\) is
\[
y_{i}
\begin{cases}
    Y_{1i} & b_{i} = 0\\
    Y_{2i} & b_{i} = 1
\end{cases}
\]

## Simulation

We demonstrate `bmixlm` with a simulated data set.
```{r}
## Generate 7 uniformly distributed covariates
set.seed(31)
d <- as.data.frame(matrix(runif(1000*7),1000,7))
colnames(d) <- letters[seq_along(d)]
```

Generate the response for the first component from covariates a, b and
c, and the response for the second from c, d and e.
```{r}
## Generate responses from the two models
sigma <- c(0.4,0.6)
beta1 <- c(runif(1,-0.4,0.4),rnorm(3))
y1 <- model.matrix(~a+b+c,d)%*%beta1+rnorm(nrow(d),0,sigma[1])
beta2 <- c(runif(1,-0.4,0.4),rnorm(3))
y2 <- model.matrix(~c+d+e,d)%*%beta2+rnorm(nrow(d),0,sigma[2])
```

The observed response is a mixture of the two responses, with the
probability of component membership determined by a probit model
dependent upon covariates f and g

```{r}
## Draw the observed response from one model or the other
betap <- c(0,1,-1)
p <- pnorm(model.matrix(~f+g,d)%*%betap)
b <- rbinom(nrow(d),1,p)
d$y <- ifelse(b==0,y1,y2)
```

In this simulated example the two components of the mixture are not
particularly distinct
```{r}
## Show the two components and the mixture
library(ggplot2)
ggplot(data.frame(comp=factor(c(b+1,rep("mixture",nrow(d)))),
                  y=c(d$y,d$y)),
       aes(x=y,group=comp,colour=comp))+
  geom_density()
```


## Fitting

Fit the model that correctly assumes that the response in the first
component is dependent upon covariates a, b and c, the response in the
second is dependent upon covariates c, d and e and the probability of
the observation is drawn from one component or the other is dependent
upon covariates f and g. The model is fit by Gibbs sampling and we
generate an initial sample of 100 draws
```{r}
## Draw a burnin sample
library(bmixlm)
fit <- bmixlm(y~a+b+c,y~c+d+e,~f+g,data=d,nsamp=100)
```
and examine traceplots to determine if the burn-in is sufficient
```{r}
## Traceplots
plot(fit,which="comp1")
plot(fit,which="comp2")
plot(fit,which="probit")
plot(fit,which="error")
```

Update the fit to draw a larger sample. Unless otherwise specified,
sampling starts from the last sample drawn in the previous fit

```{r}
## Draw a larger sample
fit <- update(fit,nsamp=2000)
summary(fit)
```

Again examine traceplots
```{r}
## Traceplots
plot(fit,which="comp1")
plot(fit,which="comp2")
plot(fit,which="probit")
plot(fit,which="error")
```

Calculate a full range of fitted values
```{r}
## Fitted values, residuals, and probability of component membership
d.pr <- predictAll(fit)
head(d.pr)
```
This yields

* `y` the observed response from the prediction data
* `y1` the prediction for a response from the first component
* `y2` the prediction for a response from the second component
* `r1` the residual if the response was from the first component
* `r2` the residual if the response was from the second component
* `p` the probability the response was drawn from the second component
  conditional on only the covariates
* `q` the probability the response was drawn from the second component
  conditional on the covariates and the response itself.

Plot residuals versus fitted values for both components, colour coded
by the probability of component membership
```{r}
## Residuals vs fitted values
d.rf <- data.frame(comp=rep(1:2,each=nrow(d.pr)),
                   fitted=c(d.pr$y1,d.pr$y2),
                   residual=c(d.pr$r1,d.pr$r2),
                   prob=c(1-d.pr$q,d.pr$q))
library(ggplot2)
ggplot(d.rf[order(d.rf$prob),],
       aes(x=fitted,y=residual,colour=prob)) +
  geom_point(size=1)+
  facet_wrap(~comp,ncol=1)
```

Show the estimated decomposition into components
```{r}
## Decompose response into components
cl <- classify(fit)
library(ggplot2)
ggplot(data.frame(comp=factor(c(ifelse(cl$q < 0.5,"1","2"),
                                rep("mixture",nrow(d)))),
                  y=rep(d$y,2)),
       aes(x=y,group=comp,colour=comp))+
  geom_density()
```



## Validation

Pairwise plots of the samples for each set of model coefficients, with the
posterior mean shown in orange and the known true value red
```{r}
## Compare samples to known parameter values
pairs(rbind(coef(fit,which="comp1",type="samples"),
            coef(fit,which="comp1",type="mean"),
            beta1),pch=16,cex=rep(c(0.7,1),c(fit$nsamp,2)),
      col=rep(c(alpha("black",0.05),alpha("orange",1),alpha("red",1)),c(fit$nsamp,1,1)),
      main="comp1")
pairs(rbind(coef(fit,which="comp2",type="samples"),
            coef(fit,which="comp2",type="mean"),
            beta2),pch=16,cex=rep(c(0.7,1),c(fit$nsamp,2)),
      col=rep(c(alpha("black",0.05),alpha("orange",1),alpha("red",1)),c(fit$nsamp,1,1)),
      main="comp2")
pairs(rbind(coef(fit,which="probit",type="samples"),
            coef(fit,which="probit",type="mean"),
            betap),pch=16,cex=rep(c(0.7,1),c(fit$nsamp,2)),
      col=rep(c(alpha("black",0.05),alpha("orange",1),alpha("red",1)),c(fit$nsamp,1,1)),
      main="probit")
pairs(rbind(coef(fit,which="error",type="samples"),
            coef(fit,which="error",type="mean"),
            sigma),pch=16,cex=rep(c(0.7,1),c(fit$nsamp,2)),
      col=rep(c(alpha("black",0.05),alpha("orange",1),alpha("red",1)),c(fit$nsamp,1,1)),
      main="error")
```

Compare the distribution of predicted probabilities of component
membership by the known membership
```{r}
## Compare estimated and known component membership
library(ggplot2)
ggplot(data.frame(comp=factor(b+1),prob=d.pr$q),
       aes(x=prob,group=comp,colour=comp))+
  geom_density()+
  xlab("Pr comp 2")
```


## Component Relabelling

It is possible to fit a simple two component mixture model by
specifying no covariates for any components of the model.
```{r}
## Fit a simple two component mixture
fit0 <- bmixlm(y~1,y~1,~1,data=d,nsamp=100)
fit0 <- update(fit0,nsamp=2000)
summary(fit0)
```
Models in which the linear models for the two components depend on
identical sets of covariates must be interpreted with care.  When there
is nothing to distinguish the models for the two components, the two
components become interchangeable and the model is subject to the
relabelling problem common to MCMC treatments of simple mixture models.


## Simulation

Simulate new observations from the posterior predictive distribution
```{r}
## Simulate from posterior predictive distribution
ys <- simulate(fit,nsim=500)
```

Plot the original observations against the 95% credible interval
generated from the posterior predictive distribution, coloured by the
certainty of component membership
```{r}
## Show observations against simulations
cl <- classify(fit)
d.pr <- as.data.frame(t(apply(ys,1,quantile,prob=c(0.025,0.5,0.975))))
d.pr <-cbind(d.pr,y=d$y,x=order(order(d.pr$`50%`)),cert=pmax(cl$q,1-cl$q))
library(ggplot2)
ggplot(d.pr,aes(x=x,y=y,ymin=`2.5%`,ymax=`97.5%`,colour=cert))+
  geom_ribbon(col="grey80",fill="grey80")+
  geom_point(size=1)
```

The equivalent plot for the simple two component mixture
model suggests the simpler model does not fit
```{r}
## Observations against simulations for simple mixture
ys <- simulate(fit0,nsim=500)
cl <- classify(fit0)
d.pr <- as.data.frame(t(apply(ys,1,quantile,prob=c(0.025,0.5,0.975))))
d.pr <-cbind(d.pr,y=d$y,x=order(order(d.pr$`50%`)),cert=pmax(cl$q,1-cl$q))
library(ggplot2)
ggplot(d.pr,aes(x=x,y=y,ymin=`2.5%`,ymax=`97.5%`,colour=cert))+
  geom_ribbon(col="grey80",fill="grey80")+
  geom_point(size=1)
```
